Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-06-29 demystify.tex
updates

\paragraph{How to help your LOR writer} Your writer can ask you to provide additional information to help them write your letter.  For example, they might ask about GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and can complement that with their own perspective.  It might also be a good idea to provide them with such information even when they do not ask, especially if you have not interacted with them much or have not done much research with them.

\paragraph{Self-written Letters} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have little value and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter).

@@

\paragraph{Response Deadlines} Accepted students are usually given a deadline to make decision on their offers, often around around \textbf{April 15th} in the US. This is a standard deadline for PhD programs due to the Council of Graduate Schools (CGS) Resolution (\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{see more). After this date, CS programs can gauge how many slots remain unfilled.

\paragraph{Waitlist} Most CS programs have a limited number of slots for PhD students, and thus put many students on a waitlist.  If accepted students decline the offer, then offers are sent to students on the waitlist. So if you see people getting accepted, that does not mean you are out yet.

\paragraph{Rejection Letters} Typically, schools start sending out rejection letters to remaining applicants after they have finalized their admissions decisions. Thus, rejection letters are often sent out late (e.g., after April 15th or even much later).

@@

For each application, the Mason application system compiles a single PDF including a summary, transcripts, test scores, LoRs, a CV, SOP, and writing samples. I usually start with the \emph{summary}, checking for low GPAs or test scores below the university minimum (\autoref{sec:standard-tests}). I then skim the \emph{transcripts} for low grades in relevant courses, noting issues like ``many low grades in main courses or unknown international school with good GPA''.  These are not as important as LORs or SOP but I read them first because the review system has questions about them (e.g., ``is GPA good? is IELTS sufficient?'')

@@

Finally, I make a decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship. The adcom chair then reviews all evaluations, looking at entered notes, and ask reviewers to discuss and resolve discrepancies to reach a consensus. Typically there is no involvement from the adcom chair, department chair, or anyone else (i.e., the decision is made by the reviewers).

@@

\begin{commentbox}[DK:]
  I was amused to read this as if I recall correctly, you never called me by my first name when you were at UNM. You always called me Prof. And, many times, I would jokingly call you back as Prof. Vu.
  \tcblower
  \textbf{Vu}: Yes, for some reason I enjoy addressing you as ``Prof.'' (without appending a last or first name).  The use of Prof. Vu may have foreshadowed my future in academia.
\end{commentbox}

@@

Here are the typical steps: (i) a faculty suspecting a cheating case will report it to the Office of Academic Integrity (OAI) at the university---the report often has supporting evidence and suggested penalty (e.g., a failing grade);  (ii)  OAI will take over and investigate the case; and (iii) OAI will make the final decision.  It is important to note that after receiving the report from your prof., OAI \emph{completely} takes over and makes decisions.  This means begging your professor will not help because they simply are no longer involved in the case.

@@

    I also visit my former professors when I am in their area. I meet Thang Bui (my MS adviser) at least once a year when I come back to Harrisburg to visit my parents. When Steph was in DC for a meeting, I invited her to Mason to give a talk on her work. I have also collaborated with them after I graduated e.g., I am working with Deepak on an NSF proposal on invariant generation.

>>>>

\paragraph{Self-written Letters} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have little value and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Moreover, experienced reviewers can easily spot (or suspect) such letters, which can be a \red{red flag}.

\paragraph{How to help your LOR writer} As mentioned above, do not write your own letter. Instead, you can help your LOR writer by providing details or unique things about yourself. For example, let them know about your GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and can complement that with their own perspective.
Sometimes your writer will explicit ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

@@

\paragraph{Response Deadlines} Accepted students are usually given a deadline to make decision on their offers, often around around \textbf{April 15th} in the US. This is a standard deadline for PhD programs due to the Council of Graduate Schools (CGS) Resolution (\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{see more}). After this date, CS programs can gauge how many slots remain unfilled.

\paragraph{Waitlist} Most CS programs have a limited number of slots for PhD students, and thus put many good students on a waitlist.  If accepted students decline the offer, then offers are sent to students on the waitlist. So if you see people getting accepted, that does not mean you are out yet.

\paragraph{Rejection Letters} Typically, schools start sending out rejection letters to remaining applicants after they have finalized their admissions decisions. Thus, rejection letters are often sent out late (e.g., after April 15th or even much later). Not much you can do here, you can try to contact the school to ask about your status, but they might not reply, they might say they are still reviewing applications, or give you inaccurate information (e.g., you will hear in two weeks -- which might not happen). In short, you just have to wait.

@@

For each application, the Mason application system compiles a single PDF file, which consists of a summary, transcripts, test scores, LoRs, a CV, SOP, and writing samples. I usually start with the \emph{summary}, checking for low GPAs or test scores below the university minimum (\autoref{sec:standard-tests}). I then skim the \emph{transcripts} for low grades in relevant courses, noting issues like ``many low grades in main courses or unknown international school with good GPA''.  These are not as important as LORs or SOP but I read them first because the review system has questions about them (e.g., ``is GPA good? is IELTS sufficient?'')

@@

Finally, I make a decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship. The adcom chair then reviews all evaluations, looking at entered notes, and ask reviewers to discuss and resolve discrepancies to reach a consensus (e.g., a reviewer wants to accept but the other wants to reject). Typically, the decision is made entirely by the reviewers; there is no involvement from the adcom chair, department chair, or others.

@@

\begin{commentbox}[DK:]
  I was amused to read this as if I recall correctly, you never called me by my first name when you were at UNM. You always called me Prof. And, many times, I would jokingly call you back as Prof. Vu.
  \tcblower
  \textbf{Vu}: Yes, for some reason I enjoy calling you ``Professor'' (without appending a last or first name).  The use of Prof. Vu may have foreshadowed my future in academia.
\end{commentbox}

@@

Here are the typical steps: (i) a faculty suspecting a cheating case will report it to the Office of Academic Integrity (OAI) at the university---the report often has supporting evidence and suggested penalty (e.g., a failing grade);  (ii)  OAI will completely take over and investigate the case independently (the prof. is not involved); and (iii) OAI will make the final decision.  It is important to note that after receiving the report from your prof., OAI \emph{completely} takes over and makes its own decision.  This means begging your professor will not help because they simply are no longer involved in the case and cannot do much.

@@

    I also visit my former professors when I am in their area. I meet Thang Bui (my MS adviser) at least once a year when I come back to Harrisburg to visit my parents. When Steph was in DC for a meeting, I invited her to to give a research talk at Mason. I have also collaborated with them after I graduated e.g., I am working with Deepak on an NSF proposal on invariant generation.
